Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-29 paper.qmd
paper updates from yesterday

@@

@@

Despite these results, it is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been apparent in existing uncertainty visualisation experiments. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. So, if we cannot see any difference to uncertainty information when it coems to value abstraction, *when* is uncertainty information different?

@@

There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust. These papers often try to ask a question that should utilise both the uncertainty and signal in the response. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].  
- @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot. Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty. The experiment set up is shown in @fig-right. The participants answers were then compared to a ground truth where the correct plot had a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.  
- @Padilla2017 made several confusing assumption when determining the ground truth of their experiments that were assessing storm visualisations. Experiment 2 assumed that the participants should provide responses that indicated they perceived the intensity of the storm to be independent of the uncertainty distribution of its location. This assumption seems counterintuitive as best, however experiment 3 asked participants to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path" for which they *were* supposed to incorperate uncertainty information and not assume they were independent. The authors seemed to be unaware of these confounding mistakes in their assumptions of how the participants were supposed to utilise the uncertainty information.  
- @Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty. Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.  
- @Correll2014 asked participants to extract the mean and variance from four uncertainty visualisations, bar charts with error bars, box plots, gradient plots, and violin plots (that were adjusted to make sure the mathematical information they were showing was the same) from two side by side distributions. Participants were also asked to answer a question that was supposed to incorperate both the signal and the noise, such as "How likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favor of A to 7=Outcome will be most in favor of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorperate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly. 
- @Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.

@@

It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, except uncertainty visualisation wants to achieve this within a single plot.

@@

Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error. This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] . The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap. Even though @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result. Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated. Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments. This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.

![](pixelmap.png){fig-pixelmap}

A final method to perform signal supression is simply to visualise the data if it is available. An example of this is shown in @fig-census for racial distributions spatially in america, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/). This map shows the typical causes of uncertainty in a spatial model, i.e. regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc, but it avoids the need to create a visualisation with a specific signal in mind. This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging. In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself. The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out. The raw data can also be used in a similar method to the HOPs plot, where a statistic of importance is implied by the visualisation, but the signal is supressed using the data itself instead of resampling methods. This method is used by the New York Times with their class mobility plot shown in fig, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html). This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through. @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the lineup protocol was implemented.

![](censusdotmap.png){fig-census}
![](classmobility.png){fig-census}

>>>>

- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.

@@

- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming. While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.

@@

Despite these results, it is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been apparent in existing uncertainty visualisation experiments. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc. There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field. Especially because it seems as though (at least in the case of visual inference) they *are* no different from any other value. So, if we cannot see any difference to uncertainty information when it comes to value abstraction, *when* is uncertainty information different?

@@

There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust. These papers often try to ask a question that should utilise both the uncertainty and signal in the response. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].  
- @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot. Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty. The experiment set up is shown in @fig-right. The participants answers were then compared to a ground truth where the correct plot had a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.  
- @Padilla2017 made several confusing assumption when determining the ground truth of their experiments that were assessing storm visualisations. Experiment 2 assumed that the participants should provide responses that indicated they perceived the intensity of the storm to be independent of the uncertainty distribution of its location. This assumption seems counterintuitive as best, however experiment 3 asked participants to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path" for which they *were* supposed to incorperate uncertainty information and not assume they were independent. The authors seemed to be unaware of these confounding mistakes in their assumptions of how the participants were supposed to utilise the uncertainty information.  
- @Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty. Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.  
- @Correll2014 asked participants to extract the mean and variance from four uncertainty visualisations, bar charts with error bars, box plots, gradient plots, and violin plots (that were adjusted to make sure the mathematical information they were showing was the same) from two side by side distributions. Participants were also asked to answer a question that was supposed to incorperate both the signal and the noise, such as "How likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favor of A to 7=Outcome will be most in favor of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorperate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly. 
- @Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location. Their payment scheme, which paid out $0.10 for a correct choice (i.e. staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didnt burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%. The authors failled to recognise that this "complicated decision making task" boilled down to a simple value extraction problem, which the text made easiest to extract. The results from the experiment indicated that many of the participants also failled to recognise the obvious and simply strategy.

- @Blenkinsop2000 tailored their questions according to the visualisation shown, which implies an understanding that different visualisations contain different information. Despite this the authors still compared the visualisations according to which were "useful for uncertainty visualisation".

- @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random . However the participants failing to identify the signal (something that would be expected of an uncertainty plot) was seen as a sign the task was too complicated. While the task in this experiment *was* certainty far too complicated (and poorly communicated) this discussion indicates the authors did not consider the signal suppression to be the goal of the visualisation, and rather an annoying distraction.

- @Boone2018

- @wu2023rational attempted to remedy this issue caused by designing a "rational agent benchmark" such that participants in an experiment are compared to a rational agent. This suggestion is an insufficient band-aid solution that is built upon the same invalid assumptions as the rest of the literature, and simply digs the field deeper into these problems, rather than solve them.

- @wu2023rational almost seems to recognise the problem of decision making experiments with payment schemes or correct/incorrect answers, but instead of understanding this boils these tasks down to value extraction, they suggest participants should be compared to a "rational agent". This rational agent acts as a benchmark, in the same way a ground truth statistic does for value extraction experiments, except the authors don't seem to realise that is what they have done. This paper highlights many decision making tasks are simply a value extraction experiment without realising it, this paper included.

@@

It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

![Difference between the null plot vs the uncertainty plot.](hypvs.jpeg){#fig-hypvs}

@@

Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error. This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] . The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html). Even though @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result. Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated. Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments. This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing. These sampling methods can show the messiness of the data that sits behind a model.

![](pixelmap.png){fig-pixelmap}
![](classmobility.png){fig-nyt}

A final method to perform signal supression is simply to visualise the data if it is available. An example of this is shown in @fig-census for racial distributions spatially in america, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/). This map shows the typical causes of uncertainty in a spatial model, i.e. regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc, but it avoids the need to create a visualisation with a specific signal in mind. This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging. In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself. The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out. The raw data can also be used in a similar method to the HOPs plot, where a statistic of importance is implied by the visualisation, but the signal is supressed using the data itself instead of resampling methods. This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through. @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the lineup protocol was implemented.

![](censusdotmap.png){fig-census}
